  ---
title: "Exploratory Data Analysis"
author: "Logan Herrmeyer"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:  
    keep_md: true
    toc: true
    toc_float: true
    fig_height: 6
    fig_width: 12
    fig_align: 'center'
---

```{r, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## Reading Notes

Exploratory data analysis is exploring data in a systematic way.

It is a time to investigate every idea that occurs to you.

The goal is to develop an understanding of the data.

It is a state of mind and not a formal process.

Categorical variables can only take a small set of values.

A continuous variable can take an infinite set of values.

Graphs can suggest more questions that we can ask.

Variation is the tendency of a variable to change from measurement to measurement.

Histograms and bar graphs are good for visualizing the distribution of a variable.

Covariation is two variables varying together, the best way to find it is to visualize the relationship on a graph.

Boxplots are also good for showing distribution.

Patterns are useful to reveal covariation.

Models are tools for extracting patterns from data.

## EDA Example

The code below is an example of the EDA process using the `starwars` data from the `tidyverse` package. (Make sure you have the `tidyverse` package installed!)

Run the code line-by-line and look at the output. Add a comment to each line of code that explains what it does/what insights it provides.

```{r}
library(tidyverse) # Import the tidyverse library

dim(starwars) # Show the dimensions (rows & cols) of the data
colnames(starwars) # Get column names of the dataset

head(starwars) # Show the first 6 cols of the dataset
glimpse(starwars) # Gives some basic information about the dataset, such as number of rows and columns, column names, column datatype, and first data points for each column.

?starwars # Gives a text description of the dataset

starwars %>% count(species) # Count the number of species observations in the dataset

mean(starwars$height) # Get the mean height in the dataset
mean(starwars$height, na.rm = TRUE) # Get the mean height in the dataset, ignoring NA (not available) values

summary(starwars$height) # Show quartiles and mean of the column

cm_to_ft <- function(cm){ # Define a function to convert centimeters to feet
  ft = cm / 30.48
  return(ft)
}

starwars_2 <- starwars %>% mutate(height_ft = cm_to_ft(height)) # Create a new dataset but with an added column of height in feet

dim(starwars_2) # Show the dimensions of the new dataset
colnames(starwars_2) # Show the column names of the dataset

summary(starwars_2$height_ft) # Show a summary (quartiles and mean) of the height in feet from the dataset

ggplot(starwars_2, aes(height_ft)) + 
  geom_histogram() # Show a histogram of the heights in feet

starwars_2 %>% filter(height_ft == max(height_ft)) # Show data from the dataset where height is only the max (show the tallest data point)
starwars_2 %>% filter(height_ft == max(height_ft, na.rm = TRUE)) #Show data from the dataset where height is only the max (show the tallest data point)

# https://starwars.fandom.com/wiki/Yarael_Poof
```

## EDA Practice

Continue exploring the `starwars` data to gain additional insights, using [R4DS: Chapter 7](https://r4ds.had.co.nz/exploratory-data-analysis.html) as a guide.

It is ok if you don't understand the all code in Chapter 7. (That is what we'll be learning the next two weeks!) If writing your own code is a struggle, try the "copy, paste, and tweak" method.

```{r}
library(tidyverse)
library(ggplot2)
glimpse(starwars)
plot(height ~ mass, data=starwars) # Are height and mass related?
ggplot(data = starwars, aes(x=birth_year)) + geom_histogram() # What is the distribution of birth years
ggplot(data = starwars, aes(x=height)) + geom_histogram() # What is the distribution of heights?
ggplot(data = starwars, aes(x=mass)) + geom_histogram() # What is the distribution of mass?
(starwars %>% filter(mass == max(mass, na.rm = TRUE)))["name"] # What character is the outlier with a large mass?
ggplot(data = starwars, mapping = aes(x = species, y = mass), na.rm = TRUE) + geom_boxplot() # How does mass vary between species?
ggplot(data = starwars, mapping = aes(x = species, y = height), na.rm = TRUE) + geom_boxplot() # How does height vary between species?
ggplot(data = starwars, mapping = aes(x = gender, y = height), na.rm = TRUE) + geom_boxplot() # How does mass vary between gender?
```